---
title: "Hands-on Exercise 10"
subtitle: "Information Dashboard Design: R methods"
date: March 21, 2024
date-modified: "last-modified"
format:
  html:
    toc: true
    number-sections: true
    code-line-numbers: true

execute:
  eval: true
  echo: true
  warning: false
---

## **Overview**

In this exercise I will:

-   create bullet chart by using ggplot2,

-   create sparklines by using ggplot2 ,

-   build industry standard dashboard by using R Shiny.

## **Loading libraries**

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
reactablefmtr, gt, gtExtras, tidyverse, svglite)
```

## The data set

### **Importing data**

```{r}
coffeechain <- read_rds("data/rds/CoffeeChain.rds")

```

### Data preparation

Aggregate Sales and Budgeted Sales at the Product level.

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()
```

### **Bullet chart in ggplot2**

```{r}
ggplot(product, aes(Product, current)) + 
  geom_col(aes(Product, max(target) * 1.01),
           fill="grey85", width=0.85) +
  geom_col(aes(Product, target * 0.75),
           fill="grey60", width=0.85) +
  geom_col(aes(Product, target * 0.5),
           fill="grey50", width=0.85) +
  geom_col(aes(Product, current), 
           width=0.35,
           fill = "black") + 
  geom_errorbar(aes(y = target,
                    x = Product, 
                    ymin = target,
                    ymax= target), 
                width = .4,
                colour = "red",
                size = 1) +
  coord_flip()
```

## **Plotting sparklines using ggplot2**

### Prepare data

```{r}
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

Compute the minimum, maximum and end of the month sales:

```{r}
mins <- group_by(sales_report, Product) %>% 
  slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>% 
  slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>% 
  filter(Month == max(Month))
```

Compute the 25 and 75 quantiles:

```{r}
quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales, 
                              0.25),
            quart2 = quantile(Sales, 
                              0.75)) %>%
  right_join(sales_report)
```

### Sparklines in ggplot2

```{r}
ggplot(sales_report, aes(x=Month, y=Sales)) + 
  facet_grid(Product ~ ., scales = "free_y") + 
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2), 
              fill = 'grey90') +
  geom_line(size=0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) + 
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title=element_blank(), axis.text.y = element_blank(), 
        axis.ticks = element_blank(), strip.text = element_blank())
```

## **Static Information Dashboard Design: `gt` and `gtExtras` methods**

### **Plotting a simple bullet chart**

```{r}
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current, 
              target = target, 
              width = 60,
              palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

## **sparklines: gtExtras method**

Prepare data:

```{r}
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate (Month = month(Date, 
                        label = TRUE, 
                        abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()
```

It is important to note that one of the requirement of gtExtras functions is that almost exclusively they require you to pass data.frame with list columns. In view of this, code chunk below will be used to convert the report data.frame into list columns.

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")
```

### Plotting coffee chain sales report

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop") %>%
   gt() %>%
   gt_plt_sparkline('Monthly Sales',
                    same_limit = FALSE)
```

### **Adding statistics**

Calculate summary statistics:

```{r}
report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)
```

### Combining the data frame

```{r}
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")

sales <- report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            )

sales_data = left_join(sales, spark)
```

### Plot updated data.table

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

### **Combining bullet chart and sparklines**

```{r}
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup() 

sales_data = sales_data %>%
  left_join(bullet)

sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales') %>%
  gt_plt_bullet(column = Actual, 
                target = Target, 
                width = 28,
                palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

## **Interactive Information Dashboard Design: reactable and reactablefmtr methods**

### Load package

```{r}
library(dataui)
```

### **Plotting interactive sparklines**

```{r}
#| code-fold: true
# Prepare list field
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))

# Plot sparklines
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

### Changing the page size

Default page size is 10, we can change it as follows:

```{r}
#| code-fold: true
reactable(
  report,
  defaultPageSize = 13, # Change page size
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

### **Adding points and labels**

`highlight_points` argument is used to show the minimum and maximum values points and `label` argument is used to label first and last values:

```{r}
#| code-fold: true
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)
```

### **Adding reference line**

`statline` argument is used to show the mean line:

```{r}
#| code-fold: true
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)
```

### Adding bandline

```{r}
#| code-fold: true
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)
```

### **Changing from sparkline to sparkbar**

```{r}
#| code-fold: true
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```

## **References**

-   [[R for Visual Analytics: Chapter 31: Information Dashboard Design: R methods]{.underline}](https://r4va.netlify.app/chap31)
