---
title: "In-class Exercise 1"
subtitle: "Now You See It"
date: January 13, 2024
date-modified: "last-modified"
format:
  html:
    toc: true
    number-sections: true
    code-line-numbers: false
    
execute: 
  eval: true
  echo: true
  warning: false  
---

## Loading R packages

In this hands-on exercise, two R packages will be used. They are:

-   [tidyverse](https://www.tidyverse.org/), and
-   [haven](https://haven.tidyverse.org/)

The code chunk used is as follows:

```{r}
pacman::p_load(tidyverse, haven)
```

::: {.callout-note}
`pacman::p_load(tidyverse, haven)` is calling the `pacman` package, since we have not yet loaded this package. If `pacman` package was already loaded, we can load `tidyverse` and `haven` with `p_load(tidyverse, haven)`.
:::

## Importing PISA data

The code chunk below uses [`read_sas()`](https://haven.tidyverse.org/reference/read_sas.html) of [`haven`](https://haven.tidyverse.org/reference/index.html) to import PISA data into R environment. 

```{r}
#| eval: false
stu_qqq <- read_sas("data/cy08msp_stu_qqq.sas7bdat")
```

The code chunk below filters the data where country code (CNT) is Singapore (SGP).

```{r}
#| eval: false
stu_qqq_SG <- stu_qqq %>% 
  filter(CNT == "SGP")
```

The code chunk below then saves this filtered data as an RData file. This smaller file can be used in the future. 

```{r}
#| eval: false
write_rds(stu_qqq_SG, "data/stu_qqq_SG.rds")
```

To load this file in the future, follow the code in the code chunk below. 

```{r}
stu_qqq_SG <- read_rds("data/stu_qqq_SG.rds")
```
