---
title: "Hands-on Exercise 7 (Work in Progress)"
subtitle: "Visualising and Analysing Geographic Data"
date: February 25, 2024
date-modified: "last-modified"
format:
  html:
    toc: true
    number-sections: true
    code-line-numbers: true
    
execute: 
  eval: true
  echo: true
  warning: false  
---

# **Choropleth Mapping with R**

## Overview

In this exercise I will learn how to plot functional and truthful choropleth maps by using an R package called [**tmap**](https://cran.r-project.org/web/packages/tmap/) package.

## Loading libraries

We will use the [**tmap**](https://cran.r-project.org/web/packages/tmap/) package that can create thematic maps, such as choropleths and bubble maps; and the [**sf**](https://cran.r-project.org/web/packages/sf/) package for handling geospatial data.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## Importing data

Two data set will be used to create the choropleth map. They are:

-   Master Plan 2014 Subzone Boundary (Web) (i.e. `MP14_SUBZONE_WEB_PL`) in ESRI shapefile format. It can be downloaded at [data.gov.sg](https://data.gov.sg/) This is a **geospatial data**. It consists of the **geographical boundary of Singapore at the planning subzone level**. The data is based on URA Master Plan 2014.

-   Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020 in csv format (i.e. `respopagesextod2011to2020.csv`). This is an aspatial data fie. It can be downloaded at [Department of Statistics, Singapore](https://www.singstat.gov.sg/) Although it does not contain any coordinates values, but it’s PA and SZ fields can be used as unique identifiers to geocode to `MP14_SUBZONE_WEB_PL` shapefile.

### **Importing Geospatial Data into R**

We first use the *st_read()* function of **sf** package to import `MP14_SUBZONE_WEB_PL` shapefile into R as a simple feature data frame called `mpsz`.

```{r}
mpsz <- st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```

::: callout-note
### Student's Notes

There are 323 features (rows) and 15 fields (columns).
:::

Let's examine the content and first ten records of `mpsz:`

```{r}
mpsz
```

::: callout-note
### Student's Notes

Each feature could represent a different area, like a subzone or district. Alongside spatial information, each feature has 15 additional attributes or fields. These fields contain non-spatial data associated with each feature, such as names, codes, and other identifiers.

Geometry type: MULTIPOLYGON: Each feature's geometry is stored as a MULTIPOLYGON, indicating that the spatial data for each area consists of one or more polygons. A polygon is a shape enclosed by a boundary that defines the area of the feature. MULTIPOLYGONs are used for complex shapes that might consist of several disjointed parts.

Dimension: XY This indicates that the spatial data is in two dimensions, using X and Y coordinates to represent locations on a plane. This is typical for most mapping and GIS applications.

Bounding box: xmin: 2667.538 ymin: 15748.72 xmax: 56396.44 ymax: 50256.33 The bounding box gives you the minimum and maximum coordinates of the dataset in the X (longitude) and Y (latitude) dimensions. This provides a rough idea of the spatial extent covered by your data.

Projected CRS: SVY21 The Coordinate Reference System defines how the two-dimensional, flat map relates to real places on the Earth. SVY21 is a specific type of CRS that is likely tailored for a particular region e.g. Singapore.
:::

### **Importing Attribute Data into R**

Next, we will import *respopagsex2011to2020.csv* file into RStudio and save the file into an R dataframe called *popagsex*.

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

```{r}
str(popdata)
```

::: callout-note
### Student's Notes

Data consists of 984,656 rows and 7 columns.

Recall that PA and SZ fields can be used as unique identifiers to geocode to MP14_SUBZONE_WEB_PL shapefile. It seems like

-   *PA* corresponds to *PLN_AREA_N*
-   *SZ* corresponds to *SUBZONE_N*

PA: Area / Region in SG SZ: Specific area within PA AG: Age group Sex: Gender TOD: Type of Dwelling Pop: Population for this age group for this area and TOD. Time: Year of info
:::

### **Data Preparation**

We first need to prepare a data table with year **2020** values. The data table should include the variables PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY.

-   YOUNG: age group 0 to 4 until age group 20 to 24,

-   ECONOMY ACTIVE: age group 25-29 until age group 60-64,

-   AGED: age group 65 and above,

-   TOTAL: all age group, and

-   DEPENDENCY: the ratio between young and aged against economy active group

#### Data wrangling

The following data wrangling and transformation functions will be used:

-   *pivot_wider()* of **tidyr** package, and

-   *mutate()*, *filter()*, *group_by()* and *select()* of **dplyr** package

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>% # Filter data for year 2020
  group_by(PA, SZ, AG) %>% # Group by PA SZ and AG. Not grouped: Sex , TOD. So we will see sum of age across each TOD and Gender for group. 
  summarise(`POP` = sum(`Pop`)) %>% # Allows us to see population for each of the group, adding up for all type of TOD and Gender. 
  ungroup() %>% 
  pivot_wider(names_from=AG, 
              values_from=POP) %>% # One row per combi of PA and SZ. Add columns for the Age Groups. Each value is the sum Pop for that age group. 
  mutate(YOUNG = rowSums(.[3:6]) # Create new column YOUNG that calculates sum of values in columns 3 to 6 for each row. The . notation is a placeholder that refers to the current data frame being manipulated. 
         +rowSums(.[12])) %>% # This is age group 5-9 which has weird position in column 12. 
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+ # Create new col again 
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%  # Create new col again 
mutate(`TOTAL`=rowSums(.[3:21])) %>%  # Create new col again but sum across all age groups
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`) # New col for the ratio between young and aged against economy active group
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, # Used to only select relevant cols so all the age Group cols are dropped. 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

#### Joining the attribute data and geospatial data

Before we can perform the georelational join, one extra step is required to **convert the values in PA and SZ fields to uppercase**. This is because the values of PA and SZ fields are made up of upper- and lowercase. On the other, hand the SUBZONE_N and PLN_AREA_N are in uppercase.

```{r}
# popdata2020 <- popdata2020 %>%
#   mutate_at(.vars = vars(PA, SZ),  # .vars argument specifies the variables (or columns) on which the function specified by .funs will be applied. 
#           .funs = funs(toupper)) %>% # .funs argument specifies the function(s) to apply to the selected variables. toupper function will be applied to each of the selected variables.
#   filter(`ECONOMY ACTIVE` > 0) # only include those where the value of the ECONOMY ACTIVE column is greater than 0.

# As `funs()` was deprecated in dplyr 0.8.0, we will use this
popdata2020 <- popdata2020 %>%
  mutate(across(.cols = c(PA, SZ), .fns = toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, *left_join()* of dplyr is used to join the geographical data and attribute table using planning subzone name e.g. *SUBZONE_N* and *SZ* as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

::: callout-note
### Student's Notes

There is no more *SZ* column in the dataframe *mpsz_pop2020*.

If the *popdata2020* data does not have a specified SZ that corresponds to SUBZONE_N, the values for PA, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY columns will reflect as NA.
:::

Save output to .rds:

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

### Choropleth Mapping Geospatial Data Using tmap

Two approaches can be used to prepare thematic map using *tmap*, they are:

-   Plotting a thematic map quickly by using *qtm()*.

-   Plotting highly customisable thematic map by using tmap elements.

#### Plotting a choropleth map quickly by using *qtm()*

The easiest and quickest to draw a choropleth map using **tmap** is using *qtm()*. It is concise and provides a good default visualisation in many cases.

The code chunk below will draw a cartographic standard choropleth map as shown below.

# References

-   [**R for Visual Analytics: Choropleth Mapping with R**](https://r4va.netlify.app/chap21)
