---
title: "Take-home Exercise 3 (Work in Progress)"
subtitle: "Be Weatherwise or Otherwise"
date: February 11, 2024
date-modified:  last-modified
format:
  html:
    toc: true
    number-sections: false
    code-line-numbers: false
    
execute: 
  eval: true
  echo: true
  warning: false  
---

# Overview

In this exercise, we will apply **visual interactivity** and **visualizing uncertainty** to validate the [claims](https://www.nccs.gov.sg/singapores-climate-action/impact-of-climate-change-in-singapore/) that:

-   Daily mean temperature are projected to increase by 1.4 to 4.6, and

-   The contrast between the wet months (November to January) and dry month (February and June to September) is likely to be more pronounced.

To do so, we will we use the daily temperature and rainfall records of **June** of the year 1983, 1993, 2003, 2013 and 2023 collected at the **Changi** climate station, to create an analytics-driven data visualisation. At the same time, we will apply appropriate interactive techniques to enhance the user experience in data discovery and/or visual story-telling.

# Getting started

## Load packages

First, we load packages required:

-   **readr**: for reading in CSV files

-   **dplyr**: for manipulating, concatenating dataframes

-   **purrr**: for handling lists and functional programming

-   **naniar**: for using *miss_vis()* function to check data for missing values

```{r}
pacman::p_load(readr, dplyr, purrr, naniar)
```

# Data preparation

## Import data

Next, we load the [Climate Historical Daily Records](http://www.weather.gov.sg/climate-historical-daily/) for the **Changi** weather station, for the five years 1983, 1993, 2003, 2013 and 2023. The data for each month of the five years was downloaded using a python script. The data, in the form of .CSV files, will be imported as a list of CSV files, then compiled into a single dataframe using R.

First, we read in all .CSV files into a list object of dataframes. As the .CSV files may be ended with either UTF-8 or ISO-8859-1 encoding, we will create a function to read in the .CSV files that encompasses both types of encoding.

```{r}
#| echo: false
# Define a function to read CSV files with different encodings
read_csv_files <- function(file_path, encodings = c("UTF-8", "ISO-8859-1")) {
  for (encoding in encodings) {
    tryCatch({
      return(read_csv(file_path, locale = locale(encoding = encoding)))
    }, error = function(e) {
      message(sprintf("Failed to decode %s with %s encoding: %s", file_path, encoding, e$message))
    })
  }
  stop(sprintf("Could not decode file %s with any of the provided encodings.", file_path))
}

# Define the names of the .CSV files
input_files <- list.files(path = "data/", pattern = "\\.csv$", full.names = TRUE)

# Read all CSV files into a list of dataframes
all_data <- map(input_files, read_csv_files)
```

Next, we identify columns with inconsistent data types across all the dataframes:

```{r}
# Define a function to get column types of a data frame
get_column_types <- function(df) {
  sapply(df, class)
}

# Get column types for each data frame
columnTypesList <- map(all_data, get_column_types)

# Identify columns with different types across data frames
uniqueTypes <- unique(unlist(columnTypesList))
columnsWithDifferentTypes <- list()

for (typeName in uniqueTypes) {
  columnsOfType <- names(Filter(function(x) typeName %in% x, columnTypesList[[1]]))
  for (colName in columnsOfType) {
    typesFound <- unique(sapply(columnTypesList, `[`, colName))
    if (length(typesFound) > 1) {
      columnsWithDifferentTypes[[colName]] <- typesFound
    }
  }
}

# Display columns with different types
columnsWithDifferentTypes
```

From the code chunk above, we observe that:

-   The 'Mean Wind Speed (km/h)' and 'Max Wind Speed (km/h)' columns exhibit data type inconsistencies.

    -   These columns will be uniformly converted to numeric type.

-   The columns 'Highest 30 Min Rainfall (mm)', 'Highest 60 Min Rainfall (mm)', and 'Highest 120 Min Rainfall (mm)' are type character in some dataframes, and NA type in other dataframes.

    -   This suggests that these columns are present in some dataframes but absent in others. We will address the handling of these columns later on.

We hence convert the 'Mean Wind Speed (km/h)' and 'Max Wind Speed (km/h)' columns to numeric type:

```{r}
#| warning: false
all_data <- map(all_data, ~ .x %>%
  mutate(
    `Mean Wind Speed (km/h)` = as.numeric(as.character(`Mean Wind Speed (km/h)`)),
    `Max Wind Speed (km/h)` = as.numeric(as.character(`Max Wind Speed (km/h)`))
  )
)
```

The list of dataframes is then combined into a single dataframe, *weather_data:*

```{r}
weather_data <- bind_rows(all_data)
```

## Clean data

Now that we have a single datafrmae, we first check the health of the dataframe by:

-   using *glimpse()* to look at the structure of the dataframe, data types of the columns, and some values of the dataframe,

-   using *duplicate()* to check the dataframe for any duplicated entries using *duplicate()*, and

-   using *vis_miss()* to check the state of missing values in the dataset.

::: panel-tabset
### glimpse()

```{r}
glimpse(weather_data)
```

### duplicated()

```{r}
weather_data[duplicated(weather_data),]
```

### vis_miss()

```{r}
vis_miss(weather_data)

```
:::

From the above, we observe these data quality issues and actionable insights:

-   The *glimpse* function shows three columns 'Year', 'Month', 'Day' representing the date of the record.

    -   These columns can be used to create a 'Date' column of date type that would make visualizing the data for different time periods easier.

-   The *glimpse* function shows the presence of character values **`"\u0097"`** in the columns 'Highest 30 Min Rainfall (mm)', 'Highest 60 Min Rainfall (mm)', and 'Highest 120 Min Rainfall (mm)'.

    -   These non-numeric placeholders represent missing values and should be converted to NA.

    -   These columns should also be of type numeric instead of character.

-   The *glimpse* function also shows there is a possible naming inconsistency of the 'Highest 30 Min Rainfall (mm)', 'Highest 60 Min Rainfall (mm)', and 'Highest 120 Min Rainfall (mm)' character type columns, where similar columns 'Highest 30 **min** Rainfall (mm)', 'Highest 60 **min** Rainfall (mm)', and 'Highest 120 **min** Rainfall (mm)' of numeric type exist. The *vis_miss()* function confirms this naming inconsistency.

    -   This can be resolved by merging the information in the character and double type columns.

```{r}
# First, create a new column Date
weather_data <- weather_data %>%
  mutate(Date = as.Date(paste(Year, Month, Day, sep = "-")))

# Next, convert character values "\u0097" to NA.
weather_data <- weather_data %>%
  mutate(
    `Highest 30 Min Rainfall (mm)` = na_if(`Highest 30 Min Rainfall (mm)`, "\u0097"),
    `Highest 60 Min Rainfall (mm)` = na_if(`Highest 60 Min Rainfall (mm)`, "\u0097"),
    `Highest 120 Min Rainfall (mm)` = na_if(`Highest 120 Min Rainfall (mm)`, "\u0097")
  )

# Next, convert these columns to numeric type.
weather_data <- weather_data %>%
  mutate(
    `Highest 30 Min Rainfall (mm)` = as.numeric(`Highest 30 Min Rainfall (mm)`),
    `Highest 60 Min Rainfall (mm)` = as.numeric(`Highest 60 Min Rainfall (mm)`),
    `Highest 120 Min Rainfall (mm)` = as.numeric(`Highest 120 Min Rainfall (mm)`)
  )

# Lastly, merge duplicate columns on the condition that the originally numeric column (e.g. Highest 30 min Rainfall (mm)) is not NA value.

weather_data <- weather_data %>%
  mutate(
    `Highest 30 Min Rainfall (mm)` = coalesce(`Highest 30 min Rainfall (mm)`, `Highest 30 Min Rainfall (mm)`),
    `Highest 60 Min Rainfall (mm)` = coalesce(`Highest 60 min Rainfall (mm)`, `Highest 60 Min Rainfall (mm)`),
    `Highest 120 Min Rainfall (mm)` = coalesce(`Highest 120 min Rainfall (mm)`, `Highest 120 Min Rainfall (mm)`)
  ) %>%
  # Optional: Remove the source columns if they are no longer needed
  select(
    -`Highest 30 min Rainfall (mm)`,
    -`Highest 60 min Rainfall (mm)`,
    -`Highest 120 min Rainfall (mm)`
  )

```

We then check the health of the data again:

::: panel-tabset
### glimpse()

```{r}
glimpse(weather_data)
```

### vis_miss()

```{r}
vis_miss(weather_data)
```
:::

Insights:

-   The dataset is now clean. It has 1825 rows and 14 columns of correct data types.

-   The missing values in the data for the three columns 'Highest 30 Min Rainfall (mm)', 'Highest 60 Min Rainfall (mm)', and 'Highest 120 Min Rainfall (mm)' correspond with [information that these datapoints were not collected before 2014](http://www.weather.gov.sg/wp-content/uploads/2022/06/Station_Records.pdf). Hence, we will leave the values for the years before 2014 as NA.

We can now proceed with data exploration.

# Data exploration

Use this section to explain your selection of data visualisation techniques used.

# Data visualization design

Write about the data visualisation design and interactivity principles and best practices implemented.

## Summary and conclusion

## References

-   
